Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions

### 2023-01-04 analysis/paper/shceof.Rmd
Todas las respuestas

output: 
  # bookdown::word_document2:
  #     pandoc_args: ["--lua-filter=abstract-to-meta.lua", "--wrap=preserve"]
  bookdown::pdf_document2:
      latex_engine: "xelatex"
      keep_tex: true
      citation_package: "natbib"
      template: "template.tex"
      pandoc_args: ["--lua-filter=abstract-to-meta.lua", "--wrap=preserve"]
always_allow_html: true
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
 - \usepackage{chngcntr}
 # - \usepackage{natbib}
 # - \usepackage{lineno}
 # - \linenumbers 
---

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 42,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

@@

The large-scale extratropical circulation in the Southern Hemisphere is strongly zonally symmetric, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
In this study we analyse the joint variability of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
The leading cEOF represents variability of a zonal wave 1 in the stratosphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes. 
One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere. 
Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2. 
As a result, the methodology proposed in this study provides a deeper understanding of the zonally asymmetric springtime extratropical SH circulation.

@@

The large-scale extratropical circulation in the Southern Hemisphere (SH) is strongly zonally symmetric, but its zonal departures are highly relevant for regional impacts [e.g. @hoskins2005].
They strongly modulate weather systems and regional climate through promoting longitudinally different latitudinal transport of heat, humidity, and momentum [@trenberth1980a; @raphael2007] and could even be related to the occurrence of high-impact climate extremes [@pezza2012].

Zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential heights or sea-level pressures at each latitude [e.g. @vanloon1972; @trenberth1980a; @turner2017].
This approach suggests that zonal waves 1 and 3 explain almost 99% of the total variance in the annual mean 500-hPa pattern at 50ºS [@vanloon1972].
@trenberth1985 concluded that wave 3 plays a role in the development of blocking events.
In addition, previous works have identified at extratropical and subpolar latitudes, wave-like patterns with dominant wavenumbers 3-4, also exerting distinctive regional impacts.
@raphael2007 shows that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration.

The Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope [@irving2015].
The latter makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
With this approach, @irving2015 showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.

Another extensively used approach to characterise the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation [@fogt2020].
SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast [@mo2001].
These patterns influence precipitation anomalies in South America [@mo2001].
Although these patterns are usually derived by applying EOF to temporal anomalies, @raphael2003 also applied EOF methods specifically to zonal anomalies.
@irving2016 proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition. 
More recently @goyal2022 created an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.

Some of the zonally asymmetric patterns of the SH circulation variability described previously, appear to have experienced secular changes.
For instance, @raphael2003 found that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
However, little is known yet about variability and trends of these patterns.

Patterns resulting from EOF analyses are more flexible than Fourier decomposition derived modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
Nonetheless, they are restricted to standing oscillation modes and could not properly represent propagating or phase-varying features such as zonal waves.
A single EOF can also represent a mixture of two or more physical modes.

A third methodology commonly used to describe circulation anomalies consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
Examples of this methodology are the SAM Index of @gong1999, the SH wave 3 activity index defined by @raphael2004 and the SH zonally asymmetric circulation index from @hobbs2010.
These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
These kinds of indices are generally easy to compute, but they usually capture non-stationary patterns.

An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions [cEOF; @horel1984].
This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
For instance, @krokhin2007 applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
Similarly, @gelbrecht2018 used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
To our knowledge, cEOF analysis has not been applied in the spatial domain to capture the phase-varying nature of planetary waves in the atmosphere.

The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using cEOF, which allow to describe phase varying planetary waves with variable amplitudes along a latitude circle.
In addition, it is proposed to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.

We restrict this paper to the spring September-October-November (SON) trimester because during this season both the tropical teleconnections over South America [@cazes-boezio2003] and the SH stratosphere-troposphere interactions [@lim2018] are at their maximum amplitude.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979 -- 2019.
Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to describe long-term trends.
We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and we computed horizontal wave activity fluxes following @plumb1985.
Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2\degree and 2.5\degree horizontal resolution respectively.
The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to the present.

@@

The amplitude of the zonal waves was obtained through the Fourier transform of the spatial field at each latitude circle.
For the analysis of wave 1, we computed its amplitude and phase by averaging (area-weighted) the data for each variable and each SON between `r knitr::combine_words(LatLabel(o3wave_lats))`, and then extracting the wave-1 component of the Fourier spectrum.
We chose this latitude band because it is wide enough to capture most of the relevant anomalies of SH mid-latitudes.

@@

(ref:eof-naive-cap) Spatial patterns of the four leading EOFs of SON zonal anomalies of geopotential height at 50 hPa south of 20\degree S for the 1979 -- 2019 period (arbitrary units).

@@

In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerate) EOFs representing similar patterns but shifted in phase [@horel1984].
Figure \@ref(fig:eof-naive) shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with higher wavenumber and three centres of action shifted by 1/4 wavelength (90\degree in frequency space).

To describe the phase-varying nature of these two waves, the simplest solution is to combine each pair of EOFs into indices of amplitude and phase.
So, for instance, the amplitude of the wave 1-like EOF could be measured as $\sqrt{\mathrm{PC1}^2 + \mathrm{PC2}^2}$ and its phase as $\tan^{-1} \left ( \frac{\mathrm{PC2}}{\mathrm{PC1}} \right )$ (where $\mathrm{PC1}$ and $\mathrm{PC2}$ are the time series associated with each EOF).
However, this rests upon visual inspection of the spatial patterns and only works properly if both phases appear cleanly in different EOFs, which is not guaranteed by construction.
In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200hPa geopotential heihgt zonal anomalies (not shown).

A better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis (Horel, 1984). 
Each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction.
The actual field reconstructed by each cEOF is the linear combination of the two spatial fields weighted by its respective time series. 
This is analogous to how any sine wave can be constructed by the sum of a sine wave and cosine wave with different amplitude but constant phase. 
This means that cEOFs naturally represent phase-varying wave-like patterns that change location as well as amplitude.

Figure \@ref(fig:ceofs-1) a.1 shows the two spatial patterns of the cEOF corresponding to XXX. 
For each subplot one of the cEOFs is plotted with shaded contours (which represents the 0º phase) and the other with black contours (which represents the 90º phase). 
This cEOF is very similar to the two leading EOFs shown in Figure \@ref(fig:eof-naive) and represents a zonal wave 1 pattern. 
For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero.
Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series will be positive and the 0º phase time series will be zero.
The intermediate phases will have non zero values in both time series.
The two phases described are not special and the same cEOF could be represented by any linear combinations of the phases shown in Figure \@ref(fig:ceofs-1) a.1. 
This parallels traditional EOFs, which are defined up to an additive constant. 
For real EOFs, this constant can be either 0 or $\pi$, corresponding to a change in sign. On the other hand, for cEOFs, it can be any real number between 0 and $2\pi$ [@horel1984], corresponding to rotations in the complex plane.

cEOFs are computed in the same way as traditional EOFs except that the data is first augmented by computing its analytic signal. 
This is a complex number whose real part is the original series and the imaginary part is the original data shifted by 90º at each spectral frequency -- i.e. 
its Hilbert transform. Each cEOF is a set of complex-valued spatial patterns and time series in which the real (or 0º phase) and imaginary (or 90º phase) parts are the two orthogonal phases of the wave-like pattern. 
In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.

The Hilbert transform is usually understood in terms of time-varying signal.
However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.

@@

The cEOF methodology is applied to zonal anomalies of SON geopotential height south of 20\degree S separately at 50 and 200 hPa.
Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
There is a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).

Both the spatial pattern similarity and the high temporal correlation of cEOFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability.
This motivates the decision of performing complex EOF jointly between levels.
Therefore cEOFs were computed using data from both levels at the same time.
In that sense each cEOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

As mentioned before, the choice of phases is arbitrary and equally valid. 
But to make the interpretation easier, we chose the phase of each cEOF so that the either the º0 cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.

Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Column Ozone and the second cEOF was closely related to ENSO.
Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0ª cEOF1 has the maximum correlation with the zonal wave 1 of Total Column Ozone between `r knitr::combine_words(LatLabel(o3wave_lats))`.
Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the º0 cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.

@@

```{r ceofs-1, fig.cap = "Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period. The shading (contours) corresponds to 0º (90º) phase. Arbitrary units. The proportion of variance explained for each mode with respect to the zonal mean is indicated in parenthesis.", fig.width=width_column, fig.height=width_column*1.2}
gdata <- ceof[, eof[[1]]$right, by = .(season)] %>% 
  copy() %>% 
  .[, hgt := hgt/sd(abs(hgt))]

@@

  tag_facets(tag = "cr")

@@

(ref:extended-series-cap) Time series of the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa. cEOF1 (row a) and cEOF2 (row b) separated in their 0º (column 1) and 90º (column 2) phase. Dark straight line is the linear trend. Black horizontal and vertical line mark the mean value and range of each time series, respectively.

@@

The cEOF1 (Fig. \@ref(fig:ceofs-1) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
At 50 hPa the 0º cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westward shift with height.
The cEOF2 (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
This modulated amplitude is more evident at 200 hPa, which is consistent with the cEOF2 computed separatedly for 200 hPa explains a bit more variance than the cEOF2 computed separatedly for 50 hPa (`r scales::percent(ceof_splitted[, eof[[1]]$sdev, by = levs][cEOF == "cEOF2", r2][1])` vs. `r scales::percent(ceof_splitted[, eof[[1]]$sdev, by = levs][cEOF == "cEOF2", r2][2])`, respectively).
There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.

@@

In the previous section, cEOFs analysis was applied to zonal anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
In addition, regression fields were computed using the full fields of the variables in order to describe the influence of the cEOFs on the temporal anomalies.

@@

Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the 0º cEOF1 is associated with a centre located over the Ross Sea.
The 90º cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica. 
At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the 0º cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with its centre shifted slightly eastward compared with the upper-level anomalies.
The 90º cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [e.g. @fogt2020].
Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.

@@

All data used in this paper available in a version-controlled repository [@campitelli2022a]. 
It is also freely available from their respective sources:

>>>>

output: 
  bookdown::word_document2:
      pandoc_args: ["--lua-filter=abstract-to-meta.lua", "--wrap=preserve"]
  bookdown::pdf_document2:
      latex_engine: "xelatex"
      keep_tex: true
      citation_package: "natbib"
      template: "template.tex"
      pandoc_args: ["--lua-filter=abstract-to-meta.lua", "--wrap=preserve"]
always_allow_html: true
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
 - \usepackage{chngcntr}
 # - \usepackage{natbib}
 # - \usepackage{lineno}
 # - \linenumbers 
---

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 44,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

@@

The large-scale extratropical circulation in the Southern Hemisphere is much more zonally symmetric than that of the Northern Hemisphere, but its zonal departures, albeit highly relevant for regional impacts, have been less studied.
In this study we analyse the joint variability of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
The leading cEOF represents variability of a zonal wave 1 in the stratosphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes. 
One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere. 
Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2. 
As a result, the methodology proposed in this study provides a deeper understanding of the zonally asymmetric springtime extratropical SH circulation.

@@

The large-scale extratropical circulation in the Southern Hemisphere (SH) is much more zonally symmetric than that of the Northern Hemisphere, but departures from the zonal mean are associated with regional impacts [e.g. @hoskins2005].
They strongly modulate weather systems and regional climate through promoting longitudinally varying meridional transport of heat, humidity, and momentum [@trenberth1980a; @raphael2007] and could even be related to the occurrence of high-impact climate extremes [@pezza2012].

The zonally asymmetric circulation is typically described by the amplitude and phase of zonal waves obtained by Fourier decomposition of geopotential height or sea-level pressure at each latitude [e.g. @vanloon1972; @trenberth1980a; @turner2017].
This approach suggests that zonal waves 1 and 3 explain almost 99% of the total variance in the annual mean 500 hPa geopotential height zonal anomalies at 50ºS [@vanloon1972].
@trenberth1985 concluded that wave 3 plays a role in the development of blocking events.
In addition, previous works have identified wave-like patterns with dominant wavenumbers 3-4 at extratropical and subpolar latitudes with distinctive regional impacts.
@raphael2007 showed that variability in the planetary wave 3 projected onto its climatological location is associated with anomalies in the Antarctic sea-ice concentration.

Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number by computing the wave envelope [@irving2015].
The wave envelope can represent planetary waves with different amplitude at different longitudes, but lacks information about phase and wave number.
Using this method, @irving2015 showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.

Another extensively-used approach to characterise the SH tropospheric circulation anomalies, is computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation [@fogt2020].
The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitudes and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast [@mo2001].
These patterns influence precipitation anomalies in South America [@mo2001].
Although these patterns are usually derived by applying EOF to temporal anomalies, @raphael2003 also applied EOF methods specifically to zonal anomalies.
@irving2016 proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition. 
More recently @goyal2022 created an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.

Some of the zonally asymmetric patterns of the SH circulation variability described previously appear to have experienced secular changes.
For instance, @raphael2003 suggests that the amplitude of the zonal wave 1 experienced a large increase and that the zonal wave 3 experienced changes in its annual cycle between 1958 and 1996.
However, little is known yet about variability and trends of these patterns.

Patterns resulting from EOF analysis are more flexible than Fourier decomposition-derived modes in that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
Nonetheless, they are restricted to standing oscillation modes and cannot properly represent propagating or phase-varying features such as zonal waves.
A single EOF can also represent a mixture of two or more physical modes.

A third methodology commonly used to describe circulation anomalies consists on identifying particular features of interest and creating indices using simple methods such as averages and differences.
Examples of this methodology are the SAM Index of @gong1999, the SH wave 3 activity index defined by @raphael2004 and the SH zonally asymmetric circulation index from @hobbs2010.
These derived methods are grounded on other methods such as Fourier decomposition or EOF to identify the centres of action for the described phenomena and can be useful to characterise features that are not readily apparent with these methods.
These kinds of indices are generally easy to compute, but they usually don't capture non-stationary patterns.

An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions (cEOF; @horel1984).
This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
For instance, @krokhin2007 applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their relationship with teleconnection indices.
Similarly, @gelbrecht2018 applied cEOF to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
To our knowledge, cEOF analysis has not been applied in the spatial domain to capture the phase-varying nature of planetary waves in the atmosphere.

The general goal of this study is to improve the description and understanding of the zonally asymmetric extratropical SH circulation using cEOF, which can describe phase varying planetary waves with variable amplitude along a latitude circle.
In addition, we try to expand the knowledge of the simultaneous behaviour of SH asymmetric circulation in the troposphere and the stratosphere.

We restrict this paper to the September-October-November (SON) trimester because during this season both the tropical teleconnections over South America [@cazes-boezio2003] and the SH stratosphere-troposphere interactions [@lim2018] are at their maximum amplitude.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and total column ozone (TOC) at 2.5º longitude by 2.5º latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 [ERA; @era5] for the period 1979 -- 2019.
Most of our analysis is restricted to the post-satellite era to avoid confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to describe long-term trends.
We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and we computed horizontal wave activity fluxes following @plumb1985.
Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2º and 2.5º horizontal resolution, respectively.
The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available from 1979 to the present.

@@

We computed the amplitude and phase of the TOC wave 1 by averaging (area-weighted) the data of each SON between `r knitr::combine_words(LatLabel(o3wave_lats))`, and then extracting the wave-1 component of the Fourier spectrum.
We chose this latitude band because it is wide enough to capture most of the relevant anomalies of SH mid-latitudes.

@@

(ref:eof-naive-cap) Spatial patterns of the four leading EOFs of SON geopotential height zonal anomalies of at 50 hPa south of 20º S for the 1979 -- 2019 period (arbitrary units).

@@

In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerate) EOFs representing similar patterns but shifted in phase [@horel1984].
Figure \@ref(fig:eof-naive) shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20º S.
It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with higher wavenumber and three centres of action shifted by 1/4 wavelength (90º in frequency space).

To describe the phase-varying nature of these two waves, the simplest solution is to combine each pair of EOFs into indices of amplitude and phase.
So, for instance, the amplitude of the wave 1-like EOF could be measured as $\sqrt{\mathrm{PC1}^2 + \mathrm{PC2}^2}$ and its phase as $\tan^{-1} \left ( \frac{\mathrm{PC2}}{\mathrm{PC1}} \right )$ (where $\mathrm{PC1}$ and $\mathrm{PC2}$ are the time series associated with each EOF).
However, this rests on visual inspection of the spatial patterns and only works properly if both phases appear cleanly in different EOFs, which is not guaranteed by construction.
In particular, this doesn't work with the wave 1 pattern observed as the leading EOF in 200 hPa geopotential height zonal anomalies (not shown).

A better alternative for describing phase-varying waves is to use Complex Empirical Orthogonal Functions (cEOF) analysis [@horel1984]. 
Each cEOF is represented by two spatial patterns that are shifted by 1/4 wavelength by construction.
The actual field reconstructed by each cEOF is the linear combination of the two spatial fields weighted by its respective time series. 
This is analogous to how any sine wave can be constructed by the sum of a sine wave and cosine wave with different amplitude but constant phase. 
This means that cEOFs naturally represent phase-varying wave-like patterns that change location as well as amplitude.

Figure \@ref(fig:ceofs-1) a.1 shows the two spatial patterns of the cEOF of geopotential height zonal anomalies. 
One of the cEOFs is plotted with shaded contours (which represents the 0º phase) and the other with black contours (which represents the 90º phase). 
This cEOF is very similar to the two leading EOFs shown in Figure \@ref(fig:eof-naive) and represents a zonal wave 1 pattern; the 0º phase is roughtly the EOF1 and the 90º phase is roughtly the EOF2).
For instance, when the phase of the wave matches the 0º phase, then the 0º phase time series will be positive and the 90º phase time series will be zero.
Similarly, when the phase of the wave matches the 90º phase, the 90º phase time series will be positive and the 0º phase time series will be zero.
The intermediate phases will have non zero values in both time series.
The two phases described are not special and the same cEOF could be represented by any linear combinations of the phases shown in Figure \@ref(fig:ceofs-1) a.1. 
This parallels traditional EOFs, which are defined up to sign, which corresponds to a rotation in the complex plane of either 0 or $\pi$.
On the other hand cEOFs are defined up to a rotation in the complex plane of any value between 0 and $2\pi$ [@horel1984].

cEOFs are computed in the same way as traditional EOFs except that the data is first augmented by computing its analytic signal. 
This is a complex number whose real part is the original series and whose imaginary part is the original data shifted by 90º at each spectral frequency -- i.e. 
its Hilbert transform. Each cEOF is a set of complex-valued spatial patterns and time series in which the real (or 0º phase) and imaginary (or 90º phase) parts are the two orthogonal phases of the wave-like pattern. 
In this paper we use the term 0º cEOF and 90º cEOF to refer to each part of the whole cEOF.

The Hilbert transform is usually understood in terms of time-varying signal.
However, in this work we apply the Hilbert transform at each latitude circle and at each considered time (e.g. the signal only depends on longitude).
Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.

@@

First, we applied the cEOF methodology is applied to zonal anomalies of SON geopotential height south of 20º S separately at 50 and 200 hPa.
Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each cEOF across levels.
There is a high degree of correlation between the magnitude of the respective cEOF1 and cEOF2 at each level.
The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown).

Both the spatial pattern similarity and the high temporal correlation of cEOFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability.
This motivates the decision of performing cEOF jointly between levels.
Therefore cEOFs were computed using data from both levels at the same time.
In that sense each cEOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

As mentioned before, the choice of phases is arbitrary and equally valid. 
But to make the interpretation easier, we chose the phase of each cEOF so that the either the 0º cEOF or the 90º cEOF is aligned with meaningful variables in our analysis.
This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.

Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Column Ozone and the second cEOF was closely related to ENSO.
Therefore, we chose the phase of cEOF1 so that the time series corresponding to the 0º cEOF1 has the maximum correlation with the zonal wave 1 of Total Column Ozone between `r knitr::combine_words(LatLabel(o3wave_lats))`.
Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the 0º cEOF2 is minimised, which also nearly maximises the correlation with the 90º cEOF2.

@@

(ref:ceofs-1-cap) Spatial patterns for the two leading cEOFs of SON geopotential height zonal anomalies at 50 hPa and 200 hPa for the 1979 -- 2019 period. The shading (contours) corresponds to 0º (90º) phase. Arbitrary units. The proportion of variance explained for each mode with respect to the zonal mean is indicated in parenthesis.

```{r ceofs-1, fig.cap = "(ref:ceofs-1-cap)", fig.width=width_column, fig.height=width_column*1.2}
gdata <- ceof[, eof[[1]]$right, by = .(season)] %>% 
  copy() %>% 
  .[, hgt := hgt/sd(abs(hgt))]

@@

  tag_facets(tag = "rc")

@@

(ref:extended-series-cap) Time series of the two leading cEOFs of SON geopotential height zonal anomalies at 50 hPa and 200 hPa. cEOF1 (row a) and cEOF2 (row b) separated in their 0º (column 1) and 90º (column 2) phase. Dark straight line is the linear trend. Black horizontal and vertical line mark the mean value and range of each time series, respectively.

@@

The cEOF1 (Fig. \@ref(fig:ceofs-1) column 1) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
At 50 hPa the 0º cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westward shift with height.
The cEOF2 (Fig. \@ref(fig:eof-naive) column 2) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, which is consistent with the the fact that the cEOF2 computed separately for 200 hPa explains a bit more variance than the cEOF2 computed separately for 50 hPa (`r scales::percent(ceof_splitted[, eof[[1]]$sdev, by = levs][cEOF == "cEOF2", r2][1])` vs. `r scales::percent(ceof_splitted[, eof[[1]]$sdev, by = levs][cEOF == "cEOF2", r2][2])`, respectively).
This suggest that this barotropic mode represents mainly tropospheric variability.

@@

In the previous section, cEOF analysis was applied to zonal anomalies derived by removing the zonally mean values in order to isolate the main characteristics of the main zonal waves characterizing the circulation in the SH.
In this section we compute regression fields using the full fields of the variables in order to describe the influence of the cEOFs on the temporal anomalies.

@@

Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the 0º cEOF1 is associated with a positive centre located over the Ross Sea.
The 90º cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica. 
At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the 0º cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with its centre shifted slightly eastward compared with the upper-level anomalies.
The 90º cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [e.g. @fogt2020].
Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.

@@

All data used in this paper available in a Zenodo repository [@campitelli2022a] (<https://zenodo.org/record/6612429>).
Indices updated monthly and daily will be made available at http://www.cima.fcen.uba.ar/~elio.campitelli/shceof/.

It is also freely available from their respective sources:
